Recursive Solutions to First-Order Model Counting

### 2022-07-31 doc/paper/paper.tex
Initial version of a truncated section

\usetikzlibrary{fit,positioning,trees,cd}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{arrows.meta}
\newcommand{\crefrangeconjunction}{--}
\crefname{algocf}{algorithm}{Algorithms}
\Crefname{algocf}{Algorithm}{Algorithms}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand\mdoubleplus{\mathbin{+\mkern-10mu+}}
\newcommand*{\twoheadrightarrowtail}{\mathrel{\rightarrowtail\kern-1.9ex\twoheadrightarrow}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\DeclareMathOperator{\CR}{\textsc{CR}}
\DeclareMathOperator{\GDR}{\textsc{GDR}}
\DeclareMathOperator{\IE}{\textsc{IE}}
\DeclareMathOperator{\Reff}{\textsc{Ref}}
\DeclareMathOperator{\wwp}{w}
\DeclareMathOperator{\wwn}{\overline{w}}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\Imm}{Im}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\size}{\sigma}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator{\WMC}{WMC}
\DeclareMathOperator{\gr}{gr}
\DeclareMathOperator{\leftlsquigarrow}{\text{\reflectbox{$\rightsquigarrow$}}}
\SetKw{KwRet}{yield}
\SetKwProg{Fn}{Function}{:}{}
\SetKwFunction{applyAllRules}{applyAllRules}
\SetKwFunction{applyGreedyRules}{applyGreedyRules}
\SetKwFunction{applyGreedyRulesToFormulas}{applyGreedyRulesToFormulas}
\SetKwFunction{emptyy}{empty}
\SetKwFunction{get}{get}
\SetKwFunction{put}{put}
\SetKwFunction{constructDomainMap}{constructDomainMap}
\SetKwFunction{identifyRecursion}{identifyRecursion}
\SetKwFunction{generateMaps}{generateMaps}
\SetKwFunction{mergeFcgs}{mergeFcgs}
\SetKwFunction{updateCache}{updCache}
\newtheorem{assumption}{Assumption}
\newtheorem{conjecture}{Conjecture}
\newtheorem{constraint}{Constraint}
\newtheorem{fact}{Fact}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{experiment}{Experiment}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\makeatletter
\newcommand{\nosemic}{\renewcommand{\@endalgocfline}{\relax}}% Drop semi-colon ;
\newcommand{\dosemic}{\renewcommand{\@endalgocfline}{\algocf@endline}}% Reinstate semi-colon ;
\newcommand{\pushline}{\Indp}% Indent
\newcommand{\popline}{\Indm\dosemic}% Undent
\makeatother
\Crefname{algocf}{Algorithm}{Algorithms}
\Crefname{constraint}{Constraint}{Constraints}
\Crefname{experiment}{Experiment}{Experiments}
\Crefname{clause}{Clause}{Clauses}
\Crefname{diagram}{Diagram}{Diagrams}
\crefname{line}{line}{lines}

@@

\begin{abstract}
  First-order model counting (FOMC) is a computational problem that asks to
  count the models of a sentence in first-order logic. Despite being around for
  more than a decade, practical FOMC algorithms are still unable to compute
  functions as simple as a factorial. We argue that the capabilities of FOMC
  algorithms are severely limited by their inability to express arbitrary
  recursive computations. To enable arbitrary recursion, we relax the
  restrictions that typically accompany domain recursion and generalise circuits
  used to express a solution to an FOMC problem to graphs that may contain
  cycles. To this end, we enhance the most well-established (weighted) FOMC
  algorithm ForcLift with new compilation rules and an algorithm to check
  whether a recursive call is feasible. These improvements allow us to find
  efficient solutions to counting fundamental structures such as injections and
  bijections.
\end{abstract}

@@

TODO: write a short informal summary. Add the full info as an appendix and
reference it.

>>>>

\usetikzlibrary{fit,positioning,trees,cd}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{arrows.meta}
\newcommand{\crefrangeconjunction}{--}
\crefname{algocf}{algorithm}{Algorithms}
\Crefname{algocf}{Algorithm}{Algorithms}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand\mdoubleplus{\mathbin{+\mkern-10mu+}}
\newcommand*{\twoheadrightarrowtail}{\mathrel{\rightarrowtail\kern-1.9ex\twoheadrightarrow}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\DeclareMathOperator{\CR}{\textsc{CR}}
\DeclareMathOperator{\GDR}{\textsc{GDR}}
\DeclareMathOperator{\IE}{\textsc{IE}}
\DeclareMathOperator{\Reff}{\textsc{Ref}}
\DeclareMathOperator{\wwp}{w}
\DeclareMathOperator{\wwn}{\overline{w}}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\Imm}{Im}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\size}{\sigma}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator{\WMC}{WMC}
\DeclareMathOperator{\gr}{gr}
\DeclareMathOperator{\leftlsquigarrow}{\text{\reflectbox{$\rightsquigarrow$}}}
\SetKw{KwRet}{yield}
\SetKwProg{Fn}{Function}{:}{}
\SetKwFunction{applyAllRules}{applyAllRules}
\SetKwFunction{applyGreedyRules}{applyGreedyRules}
\SetKwFunction{applyGreedyRulesToFormulas}{applyGreedyRulesToFormulas}
\SetKwFunction{emptyy}{empty}
\SetKwFunction{get}{get}
\SetKwFunction{put}{put}
\SetKwFunction{constructDomainMap}{constructDomainMap}
\SetKwFunction{identifyRecursion}{r}
\SetKwFunction{generateMaps}{genMaps}
\SetKwFunction{mergeFcgs}{mergeFcgs}
\SetKwFunction{updateCache}{updCache}
\newtheorem{assumption}{Assumption}
\newtheorem{conjecture}{Conjecture}
\newtheorem{constraint}{Constraint}
\newtheorem{fact}{Fact}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{experiment}{Experiment}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\makeatletter
\newcommand{\nosemic}{\renewcommand{\@endalgocfline}{\relax}}% Drop semi-colon ;
\newcommand{\dosemic}{\renewcommand{\@endalgocfline}{\algocf@endline}}% Reinstate semi-colon ;
\newcommand{\pushline}{\Indp}% Indent
\newcommand{\popline}{\Indm\dosemic}% Undent
\makeatother
\Crefname{algocf}{Algorithm}{Algorithms}
\Crefname{constraint}{Constraint}{Constraints}
\Crefname{experiment}{Experiment}{Experiments}
\Crefname{clause}{Clause}{Clauses}
\Crefname{diagram}{Diagram}{Diagrams}
\crefname{line}{line}{lines}

@@

\begin{abstract}
  First-order model counting (FOMC) is a computational problem that asks to
  count the models of a sentence in first-order logic. Despite being around for
  more than a decade, practical FOMC algorithms are still unable to compute
  functions as simple as a factorial. We argue that the capabilities of FOMC
  algorithms are severely limited by their inability to express arbitrary
  recursive computations. To enable arbitrary recursion, we relax the
  restrictions that typically accompany domain recursion and generalise circuits
  used to express a solution to an FOMC problem to graphs that may contain
  cycles. To this end, we enhance the most well-established (weighted) FOMC
  algorithm \textsc{ForcLift} with new compilation rules and an algorithm to
  check whether a recursive call is feasible. These improvements allow us to
  find efficient solutions to counting fundamental structures such as injections
  and bijections.
\end{abstract}

@@

% TODO: will need to explain: hashing, caching, how the output of genMaps is filtered

\paragraph*{Notation.}
First, for partial functions $\alpha, \beta\colon A \pfun B$ s.t.
$\alpha|_{\dom(\alpha) \cap \dom(\beta)} = \beta|_{\dom(\alpha) \cap \dom(\beta)}$,
we write $\alpha \cup \beta$ for the unique partial function s.t.
$\alpha \cup \beta|_{\dom(\alpha)} = \alpha$, and
$\alpha \cup \beta|_{\dom(\beta)} = \beta$. Second, let $\Doms$ be a function
that maps any clause or formula to the set of domains used within. Specifically,
$\Doms(c) \coloneqq \Imm \delta_c$ for any clause $c$, and
$\Doms(\phi) \coloneqq \bigcup_{c \in \phi} \Doms(c)$ for any formula $\phi$.

\begin{algorithm}[t]
  \caption{The compilation rule for $\Reff$ nodes.}\label{alg:trycache}
  \KwIn{formula $\phi$, cache $C$}
  \KwOut{a set of chips}

  \ForEach{formula and node $(\psi, v) \in C(\#\phi)$}{
    $\rho \gets \identifyRecursion{$\phi$, $\psi$}$\;
    \lIf{$\rho \ne {\normalfont \texttt{null}}$}{\Return{$\{\, (\Reff_\rho(v), \langle\rangle) \,\}$}}
  }
  \Return{$\emptyset$}\;

  \Fn{\identifyRecursion{$\phi$, $\psi$, $\rho = \emptyset$}}{
    \lIf{$|\phi| \ne |\psi|$ {\bf or} $\#\phi \ne \#\psi$}{\Return{\normalfont\texttt{null}}}
    \lIf{$\phi = \emptyset$}{\Return{$\rho$}}
    \ForEach{clause $c \in \psi$\label{line:for1}}{
      \ForEach{clause $d \in \phi$ s.t. $\#d=\#c$\label{line:for2}}{
        \ForAll{$(\beta, \gamma) \in \generateMaps{$c$, $d$, $\rho$}$\label{line:generateMaps}}{
          $\rho' \gets \identifyRecursion{$\phi\setminus\{\, d \,\}$, $\psi\setminus\{\, c \,\}$, $\rho\cup\gamma$}$\;\label{line:recursion}
          \lIf{$\rho' \ne {\normalfont \texttt{null}}$}{\Return{$\rho'$}}
        }
      }
      \Return{\normalfont\texttt{null}}\;
    }
  }
\end{algorithm}

% Cache $C$ is used to partition all previously-encountered formulas based on
% their hash codes.
\Cref{alg:trycache} describes the compilation rule for creating $\Reff$ nodes.
For every formula $\psi$ s.t. $\#\psi = \#\phi$ that we have encountered so far,
function \texttt{r} is called to check whether a recursive call
is actually feasible. If it is, the function returns a (total) map
$\rho\colon \Doms(\psi) \to \Doms(\phi)$ that shows how $\psi$ can be
transformed into $\phi$ by replacing each domain $\Omega \in \Doms(\psi)$ with
$\rho(\Omega) \in \Doms(\phi)$. Otherwise, \texttt{r} returns
\texttt{null} to signify that $\phi$ and $\psi$ are too different for recursion
to work. This happens if $\phi$ and $\psi$ (or their subformulas explored in
recursive calls) are structurally different (i.e., the numbers of clauses or the
hash codes fail to match) or if a clause of $\psi$ cannot be paired with a
sufficiently similar clause of $\phi$.

Function \texttt{r} iterates over pairs of clauses of $\phi$ and
$\psi$ that have the same hash codes. It uses function \texttt{genMaps} to
check whether the two clauses are sufficiently similar. If so, the function
calls itself on the remaining clauses until the map
$\rho\colon \Doms(\psi) \pfun \Doms(\phi)$ becomes total, and all clauses are
successfully coupled.

Given two clauses $c \in \psi$ and $d \in \phi$, \texttt{genMaps} considers
all possible bijections\footnote{Although the number of bijections between two
  sets of cardinality $n$ is $n!$, this part of the algorithm is unlikely to
  cause performance issues for two reasons. First, in practice, $n$ is usually
  at most two or three. Second, due to the manner in which formulas are modified
  by compilation rules, if any bijection results in a successfully identified
  recursive relationship, it is almost always the identity bijection.}
$\beta\colon \Vars(c) \twoheadrightarrowtail \Vars(d)$ and calls
\texttt{constructDomainMap}, which then attempts to construct a map
$\gamma\colon \Doms(c) \to \Doms(d)$ consistent with both $\beta$ and (the as
yet partial map) $\rho\colon \Doms(\psi) \pfun \Doms(\phi)$. The \textbf{yield}
keyword in \texttt{genMaps} works as in programming languages such as C\#,
JavaScript, and Python, and lazily returns a sequence of values, computing each
element of the sequence as needed.

Diagrammatically, \texttt{constructDomainMap} attempts to find a
$\gamma\colon \Doms(c) \to \Doms(d)$ s.t.
\begin{equation}\label[diagram]{eq:commute}
  \begin{tikzcd}
    V = \Vars(c) \ar[r, tail, two heads, "\beta"] \arrow[d, swap, "\delta_c"] & \Vars(d) \ar[d, "\delta_d"] \\
    \Doms(c) \ar[r, dashed, "\gamma"] \ar[d, hookrightarrow] & \Doms(d) \ar[d, hookrightarrow] \\
    \Doms(\psi) \ar[r, swap, "\rho", "|" marking, outer sep=5pt] & \Doms(\phi).
  \end{tikzcd}
\end{equation}
commutes (and returns \texttt{null} if such a function does not exist). Indeed,
for every variable in $V = \Vars(c)$, the function returns \texttt{null} if
either the top rectangle from $V$ to $\Doms(d)$ or the outer rectangle from $V$
to $\Doms(\phi)$ fails to commute. These checks also ensure that
$\rho \cup \gamma$ is possible on \cref{line:recursion} of the algorithm, i.e.,
$\rho|_{\dom(\rho)\cap\dom(\gamma)} = \gamma|_{\dom(\rho)\cap\dom(\gamma)}$.

% \begin{example}\label{example}

%   % definitions of formulas and intro
%   As in \cref{example:first}, let $\psi \coloneqq \{\, c_1, c_2 \,\}$ be a
%   formula with clauses
%   \begin{align*}
%     c_1 &= (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}), \\
%     c_2 &= (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (X, Z) \,\}, \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Gamma \,\}).
%   \end{align*}
%   Let formula
%   $\phi \coloneqq \psi[\id, \{\, \Gamma \mapsto \Gamma', \Delta \mapsto \Delta^\bot \,\}]$
%   be just like $\psi$ but with different domains. In other words,
%   $\phi = \{\, d_1, d_2 \,\}$, where
%   \begin{align*}
%     d_1 &\coloneqq (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}),\\
%     d_2 &\coloneqq (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (X, Z) \,\}, \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Gamma' \,\}).
%   \end{align*}
%   Note that $\#\phi = \#\psi$ and assume that $(\psi, v) \in C(\#\phi)$ for some
%   node $v$. We shall see how \cref{alg:trycache} identifies that the FCG for
%   $\psi$ can be reused for $\phi$ as well.

%   Since both formulas are non-empty, the algorithm proceeds with the for-loops
%   on \cref{line:for1,line:for2,line:generateMaps}. Suppose $c = c_1$ and
%   $d = d_1$ get picked. Since both clauses have three variables, in the worst
%   case, function \texttt{genMaps} would have $3!=6$ bijections to check.
%   Suppose the identity bijection is picked first. Then
%   \texttt{constructDomainMap} is called with the following parameters:
%   \begin{itemize}
%     \item $V = \{\, X, Y, Z \,\}$,
%     \item
%           $\delta_c = \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}$,
%     \item
%           $\delta_d = \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}$,
%     \item $\beta = \id = \{\, X \mapsto X, Y \mapsto Y, Z \mapsto Z \,\}$,
%     \item $\rho = \emptyset$.
%   \end{itemize}
%   Since $\delta_c(Y) = \delta_c(Z)$, and $\delta_d(Y) = \delta_d(Z)$,
%   \texttt{constructDomainMap} returns
%   $\gamma = \{\, \Gamma \mapsto \Gamma', \Delta \mapsto \Delta^\bot \,\}$. Thus,
%   \texttt{genMaps} yields its first pair of maps $(\beta, \gamma)$ to
%   \cref{line:generateMaps}. Furthermore, this pair satisfies
%   $c[\beta, \gamma] = d$. On \cref{line:recursion}, a recursive call to
%   \texttt{r($\{\,c_2\,\}$, $\{\,d_2\,\}$, $\gamma$)} is made.

%   In this subproblem where both formulas are left with a single clause each,
%   again we have two non-empty formulas with equal hash codes. Thus
%   \texttt{genMaps} is called with $c = c_2$, $d = d_2$, and
%   $\rho = \{\, \Gamma \mapsto \Gamma', \Delta \mapsto \Delta^\bot \,\}$. Suppose
%   \cref{line:bijection} picks the identity bijection gain. Then
%   \texttt{constructDomainMap} is called with the following parameters:
%   \begin{itemize}
%     \item $V = \{\, X, Y, Z \,\}$,
%     \item
%           $\delta_c = \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Gamma \,\}$,
%     \item
%           $\delta_d = \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Gamma' \,\}$,
%     \item $\beta = \id = \{\, X \mapsto X, Y \mapsto Y, Z \mapsto Z \,\}$,
%     \item $\rho = \{\, \Gamma \mapsto \Gamma', \Delta \mapsto \Delta^\bot \,\}$.
%   \end{itemize}
%   Since $\beta$ and $\rho$ commute (as in \cref{eq:commute}), and there are no
%   new domains in $\Doms(c)$ and $\Doms(d)$, $\gamma$ exists and is equal to
%   $\rho$. Again, the returned pair $(\beta, \gamma)$ satisfies
%   $c[\beta, \gamma] = d$. \Cref{line:recursion} calls
%   \texttt{r($\emptyset$, $\emptyset$, $\rho$)}, which
%   immediately returns
%   $\rho = \{\, \Gamma \mapsto \Gamma', \Delta \mapsto \Delta^\bot \,\}$ as the
%   final answer. Therefore, one can indeed reuse an FCG for $\psi$ to compute the
%   model count of $\phi$.
% \end{example}
